### 2023-09-01 sections/2-methods.tex
more clarity changes

\subsection {Index-Building and Sharded Search}
In this work, we also introduce the method of \emph{sharding} for $k$-nearest neighbors search.
With this approach, we aim to leverage the fact that $\rho$-nearest neighbors search is typically 
much faster than $k$-nearest neighbors search.

We begin by randomly partitioning the dataset into $s$ shards, where $s$ is determined by [TODO: how did 
we decide we're going to select $s$ for this version of the paper?]. On the first shard, we perform 
$k$-nearest neighbors search using one of the four algorithms described in Section ~\ref{subsec:methods:knn-search}.
The result of this search will give us the $k$ closest neighbors to the query in the first shard. We then take the furthest of 
those $k$ neighbors and use its distance from the query as the radius for $\rho$-nearest neighbors search (see Section ~\ref{subsec:methods:rnn-search}) 
on each remaining shard. We then take the union of the results of these $\rho$-nearest neighbors searches, along with the $k$ neighbors
from the first shard, and then linear search over those points to find the actual $k$ nearest neighbors.

Note that the size of the shards is not necessarily consistent; it is often optimal to have the first shard be smaller than the others,
since the first shard is the only one for which we perform $k$-nearest neighbors search. However, making this first shard too small
creates the risk of significantly overestimating the radius, resulting in much slower $\rho$-nearest neighbors search on the remaining shards.
Finding the precise optimal ratio between the size of the first shard and the size of the remaining shards, as well 
as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the 
actual $k$-th closest neighbor, are topics for future work.

@@

We conduct $\rho$-nearest neighbors search as described in \cite{ishaq2019clustered}, but 
with the following improvement: when a cluster overlaps with the query ball, instead of  
always proceeding with both of its children, we only proceed with those children which 
also overlap with the query ball.

@@

In this section, we present four novel algorithms for $k$-nearest neighbors search: $k$-NN by Repeated $\rho$-NN, Sieve Search, and Sieve Search with Separate Centers, 
and  Greedy Sieve. 
We also have an implementation of linear search. We use a preliminary auto-tuning function to predict the variant which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant.

In these algorithms, we typically use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far and
$Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the nearest neighbors.

@@

In this algorithm, we perform $\rho$-nearest neighbors search at a radius equal to the radius of the root cluster divided by
the cardinality of the dataset, repeatedly increasing the search radius until $k$ neighbors
have been found.

@@

Once $|H| > 0$, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension of the clusters containing the points found so far. In particular, 
we increase the radius by a factor of 
\begin{equation} \argmin \left(2, \left({\frac{k}{|H|}}\right)^{\frac{1}{\mu}}\right) \label{2} \end{equation}
where $\mu$ is the local fractal dimension (abbreviated lfd in algorithm below) in the region around the query point.
Intuitively, the factor by which we increase the radius should be \emph{inversely} related to the number of points found so far. 
Additionally, when the local fractal dimension at the radius scale from the previous iteration is low, this suggests that the data 
are relatively concentrated in that region; thus, the factor of radius increase should be \emph{directly} related to the 
local fractal dimension. Once $|H| >= k$, we return the $k$ closest points.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated$\rho$-NN(\emph{root, query, k})} % give the algorithm a caption
    \label{alg:knn-by-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[4] % enter the algorithmic environment
        \STATE $H \Leftarrow$ $\emptyset$
        \STATE $m \Leftarrow$ $\frac{root.radius}{|root|}$
        \WHILE {$|H| = 0$}
            \STATE $H.push(\rho$-NN$(root, query, m)$)
            \STATE $m \Leftarrow 2m$
        \ENDWHILE
        \WHILE {$|H| < k$}
            \STATE $Q \Leftarrow \{ C: \exists p \in H \land p \in C \}$
            \STATE $\mu \Leftarrow \frac{1}{|Q|} \sum_{C \in Q} C.lfd$
            \STATE $m \Leftarrow \argmin \left(2, \left({\frac{k}{|H|}}\right)^{\frac{1}{\mu}}\right)$
            \STATE $H.push(\rho$-NN$(root, query, m)$)
        \ENDWHILE
        \STATE $H.sort()$
        \STATE Return $H[0], H[1], \cdots H[k-1]$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve Search}
\label{subsubsec:methods:knn-search:sieve}
With Sieve Search, we begin by letting $Q$ be a list initialized with the root cluster. This list will maintain the 
clusters and points which are still in contention for being one of the nearest neighbors.
While $Q$ contains at least one cluster (i.e., $Q$ is not a list of only points), we repeat the process described in 
the following paragraphs.

@@

To find this element, 
we use a process similar to the Partition algorithm used in Quicksort~\cite{10.1093/comjnl/5.1.10}, adjusted to account for the varying cardinalities of elements in $Q$. Our modified algorithm finds the smallest index $i$ such 
that all elements in $Q$ with $\delta_{max}$ closer to or equal to $Q[i]$'s $\delta_{max}$ collectively have cardinality greater
than or equal to $k$. In the process of finding this index, we alter the order of $Q$ so that all elements to left
of index $i$ have $\delta_{max}$ less than or equal to $Q[i]$'s $\delta_{max}$ and all elements to right of index $i$
have $\delta_{max}$ greater than $Q[i]$'s $\delta_{max}$. We consider $Q[i]$ to be our $q_{\tau}$, and our threshold 
$\tau$ to be $Q[i]$'s $\delta_{max}$.

@@

We complete this process until $Q$ contains only points. At this point, we 
just use [INSERT ref to quicksort partition in paper] to find the $k$th 
nearest neighbor. Since this algorithm will also elements on the correct side of 
the index of the $k$th nearest neighbor, the first $k$ elements in $Q$ are the $k$-nearest neighbors.

@@

\subsubsection{Sieve Search with Separate Centers}
\label{subsubsec:methods:knn-search:sieve2}
Sieve Search with Separate Centers is the same as Sieve Search, but with the following modification: clusters 
in $Q$ are represented twice-- once as their center and once as the rest of their points. 
Because we have that for any cluster $C$, $C.\delta \leq C.\delta_{max}$, representing the cluster 
center separately from the rest of the points causes the threshold $\tau$ to decrease more quickly, 
thus allowing us to eliminate some clusters from $Q$ earlier in the process.

@@

With Greedy Sieve, we first let $Q$ be a min queue of clusters by non-increasing $\delta_{min}$. We initialize $Q$ with the root cluster.
Let $H$ be a fixed-size ($k$) max queue of points by $\delta$. This queue starts empty.

While $H$ has fewer than $k$ clusters or $\delta$ of the top priority element in $H$ is greater 
than $\delta_{min}$ of the top priority element in $Q$, do the following:
\begin{enumerate}
\item While the top priority element $q$ in $Q$ is not a leaf, pop $q$ from the queue and push its children.
\item For each point $p \in q$, push $p$ to $H$. 
\item Pop points from $H$ until $H$ has $k$ points. 
\end{enumerate}
After this process, the points left in $H$ are the $k$ nearest neighbors of $q$.

>>>>

\subsection {Index-Building and Sharded Search}
\label{subsec:methods:index-building}
In this work, we also introduce the method of \emph{sharding} for $k$-nearest neighbors search.
With this approach, we aim to leverage the fact that $\rho$-nearest neighbors search is typically 
much faster than $k$-nearest neighbors search.

We begin by randomly partitioning the dataset into $s$ shards, where $s$ is determined based 
on some naive auto-tuning. On the first shard, we perform 
$k$-nearest neighbors search using one of the four algorithms described in Section~\ref{subsec:methods:knn-search}.
The result of this search will give us the $k$ closest neighbors to the query \emph{in the first shard}. We then take the furthest of 
those $k$ neighbors and use its distance from the query as the radius for use with $\rho$-nearest neighbors search (see Section~\ref{subsec:methods:rnn-search}) 
on each remaining shard. We then take the union of the results of these $\rho$-nearest neighbors searches, along with the $k$ neighbors
from the first shard, and then linear search over those points to find the actual $k$ nearest neighbors.

Note that the size of the shards is not necessarily consistent; it is often optimal to have the first shard be smaller than the others,
since the first shard is the only one for which we perform $k$-nearest neighbors search. However, making this first shard too small
introduces the risk of significantly overestimating the radius, resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
Finding the precise optimal ratio between the size of the first shard and the size of the remaining shards, as well 
as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the 
actual $k$-th closest neighbor, are topics for future work.

@@

We conduct $\rho$-nearest neighbors search as described in \cite{ishaq2019clustered}, but 
with the following improvement: when a cluster overlaps with the query ball, instead of  
always proceeding with both of its children, we proceed only with those children which 
also overlap with the query ball. This improvement is demonstrated in Algorithm \ref{alg:rnn}.

@@

In this section, we present four novel algorithms for exact $k$-nearest neighbors search: $k$-NN by Repeated $\rho$-NN, Sieve Search, Sieve Search with Separate Centers, 
and Greedy Sieve. 
We use a preliminary auto-tuning function to predict the variant and shard size (see Section~\ref{subsec:methods:index-building}) which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant on the first shard.

In these algorithms, we typically use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far and
$Q$ to refer to the data structure which stores the clusters (and, for Sieve and Sieve with Centers, points) which are still in contention for being one of the nearest neighbors.

@@

In this algorithm, we perform $\rho$-nearest neighbors search small search radius, repeatedly increasing the radius until $k$ neighbors
have been found.

@@

Once $|H| > 0$, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension in the region of the dataset around the query ball. 
In particular, 
we increase the radius by a factor of 
\begin{equation} \argmin \left(2, \left({\frac{k}{|H|}}\right)^{\frac{1}{\mu}}\right) \label{2} \end{equation}
where $\mu$ is the harmonic mean of the local fractal dimensions (abbreviated lfd in algorithm below) of 
each cluster which contains one of the nearest neighbors found thus far. We use the harmonic mean to 
ensure that $\mu$ is not dominated by outlier clusters with very high fractal dimension. 
Intuitively, the factor by which we increase the radius should be \emph{inversely} related to the number of points found so far. 
Additionally, when the local fractal dimension at the radius scale from the previous iteration is low, this suggests that the data 
are relatively concentrated in that region; thus, the factor of radius increase should be \emph{directly} related to the 
local fractal dimension. Once $|H| >= k$, we return the $k$ closest points.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated$\rho$-NN(\emph{root, query, k})} % give the algorithm a caption
    \label{alg:knn-by-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[4] % enter the algorithmic environment
        \STATE $H \Leftarrow$ $\emptyset$
        \STATE $m \Leftarrow$ $\frac{root.radius}{|root|}$
        \WHILE {$|H| = 0$}
            \STATE $H.push(\rho$-NN$(root, query, m)$)
            \STATE $m \Leftarrow 2m$
        \ENDWHILE
        \WHILE {$|H| < k$}
            \STATE $Q \Leftarrow \{ C: \exists p \in H \land p \in C \}$
            \STATE $\mu \Leftarrow |Q| \sum_{C \in Q} \frac{1}{C.lfd} $
            \STATE $m \Leftarrow \argmin \left(2, \left({\frac{k}{|H|}}\right)^{\frac{1}{\mu}}\right)$
            \STATE $H.push(\rho$-NN$(root, query, m)$)
        \ENDWHILE
        \STATE $H.sort()$
        \STATE Return $H[0], H[1], \cdots H[k-1]$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve Search}
\label{subsubsec:methods:knn-search:sieve}
With Sieve Search, we begin by letting $Q$ be a list containing the root cluster. 
While $Q$ contains at least one cluster (i.e., $Q$ is not a list of only points), we repeat the process described in 
the following paragraphs.

@@

To find this element, 
we use a process similar to the Partition algorithm used in Quicksort~\cite{10.1093/comjnl/5.1.10}, adjusted to account for the varying cardinalities of elements in $Q$. 
Our modified version of this algorithm (see Algorithm 4) finds the smallest index $i$ such 
that all elements in $Q$ with $\delta_{max}$ closer to or equal to $Q[i]$'s $\delta_{max}$ collectively have cardinality greater
than or equal to $k$. We consider $Q[i]$ to be our $q_{\tau}$, and our threshold 
$\tau$ to be $Q[i]$'s $\delta_{max}$.

@@

We complete this process until $Q$ contains only points. At this point, we 
just use Algorithm 4 to find the $k$th 
nearest neighbor and return all points to the left of it.

This process is described in Algorithm~\ref{alg:sieve}.

@@

\subsubsection{Sieve Search with Separate Centers}
\label{subsubsec:methods:knn-search:sieve2}
Sieve Search with Separate Centers is the same as Sieve Search, but with the following modification: clusters 
in $Q$ are represented twice-- once as their center and once as the rest of their points. 
Because we have that for any cluster $C$, $C.\delta \leq C.\delta_{max}$, representing the cluster 
center separately from the rest of the points causes the threshold $\tau$ to decrease more quickly, 
thus allowing us to eliminate some clusters from $Q$ earlier in the process. More discussion about 
when this algorithm is preferable to Sieve Search is in Section 4: Results.

@@

With Greedy Sieve, we first let $Q$ be a min queue of clusters based on $\delta_{min}$. We initialize $Q$ with the root cluster.
Let $H$ be a fixed-size ($k$) max queue of points based on $\delta$. This queue starts empty.

While $H$ has fewer than $k$ clusters or $\delta$ of the top priority element in $H$ is greater 
than $\delta_{min}$ of the top priority element in $Q$, we do the following:
\begin{enumerate}
\item While the top priority element $q$ in $Q$ is not a leaf, pop $q$ from $Q$ and push its children.
\item For each point $p \in q$, push $p$ to $H$. 
\item Pop from $H$ until $H$ has $k$ points. 
\end{enumerate}
After this process, the points left in $H$ are the $k$ nearest neighbors of $q$.

This process is described in Algorithm \ref{alg:greedy_search}.

### 2023-09-01 sections/3-datasets.tex
more clarity changes

\subsection{APOGEE-2}
\label{subsec:datasets:apogee-2}
The Sloan Digital Sky Survey (SDSS) [3] Apache Point Observatory Galaxy Evolution
Experiment (APOGEE) [4] contains near-infrared
spectra of approximately 130,000 stellar objects in the Milky
Way galaxy. Each spectral datum describes a star's flux in
Janskys ($1 Jy = 10^{-26} W \dot m^{-2} \dot Hz^{-1}$) at each of 8,575 wavelengths ranging from $\lambda$ = 1.51 µm to 1.70 µm. Thus,
Thus, each datum is a real-valued vector in $\mathbb{R}_{+}^{8575}$.

Astronomers use these near-infrared spectra to achieve a better understanding of our galaxy's history. This dataset 
is often used with Euclidean (L2) distance, cosine distance, and 1d-Wasserstein distance. As illustrated by Figure 1 in Section 2,
this dataset is a good candidate for CAKES because it exhibits low fractal dimension.

@@

\dots

@@

We benchmark on a variety of datasets from the ann-benchmarks suite~\cite{Aumller2018ANNBenchmarksAB}, 
which we list below, along with the distance functions with which they are typically used.

\begin{itemize} 
    \item DEEP1B (Cosine): ~1 billion 96-dimensional deep descriptors to 
    \item Fashion-MNIST (Euclidean): 60,000 784-dimensional vector representations of images of clothing items
    \item GIST (Euclidean): 1 million 960-dimensional image descriptors 
    \item GloVe (Cosine): ~1.2 million 25-dimensional vector representations of words
    \item Kosarak (Jaccard): ~75 thousand ~28 thousand-dimensional binary vectors representing user clickstreams
    \item MNIST (Euclidean): 60,000 784-dimensional vector representations of images of handwritten digits
    \item NYTimes (Cosine): 290,000 256-dimensional vector representations of words
    \item SIFT (Euclidean): 1 million 128-dimensional image descriptors
    \item Last.fm (Cosine): ~292, 000 65-dimensional vector representations of songs
\end{itemize}

>>>>

\subsection{APOGEE-2}
\label{subsec:datasets:apogee-2}
The Sloan Digital Sky Survey (SDSS) [3] Apache Point Observatory Galaxy Evolution
Experiment (APOGEE) [4] contains near-infrared
spectra of approximately 130,000 stellar objects in the Milky
Way galaxy. Each spectral datum describes a star's flux in
Janskys ($1 Jy = 10^{-26} W \dot m^{-2} \dot Hz^{-1}$) at each of 8,575 wavelengths ranging from $\lambda$ = 1.51 µm to 1.70 µm. 
Thus, each datum is a real-valued vector in $\mathbb{R}_{+}^{8575}$.

Astronomers use these near-infrared spectra to achieve a better understanding of our galaxy's history. This dataset 
is often used with Euclidean (L2) distance, cosine distance, and 1d-Wasserstein distance. As explained in~\cite{ishaq2019clustered},
this dataset is a good candidate for entropy-scaling search algorithms like CAKES because it exhibits low fractal dimension.

@@

@@

We benchmark on a variety of datasets from the ann-benchmarks suite~\cite{Aumller2018ANNBenchmarksAB}, 
listed below, along with the appropriate distance function:

\begin{itemize} 
    \item DEEP1B (Cosine): $\sim$1 billion 96-dimensional deep descriptors to 
    \item Fashion-MNIST (Euclidean): 60,000 784-dimensional vector representations of images of clothing items
    \item GIST (Euclidean): 1 million 960-dimensional image descriptors 
    \item GloVe (Cosine): $\sim$1.2 million 25-dimensional vector representations of words
    \item Kosarak (Jaccard): $\sim$75 thousand ~28 thousand-dimensional binary vectors representing user clickstreams
    \item MNIST (Euclidean): 60,000 784-dimensional vector representations of images of handwritten digits
    \item NYTimes (Cosine): 290,000 256-dimensional vector representations of words
    \item SIFT (Euclidean): 1 million 128-dimensional image descriptors
    \item Last.fm (Cosine): $\sim$292, 000 65-dimensional vector representations of songs
\end{itemize}
